---
title: "Xenia Tornado"
author: "Tyler Fricker"
date: "2/19/2019"
output: html_document
---

Set working directory and load packages.
```{r}
library(tidyverse)
library(lubridate)
library(sf)
library(tmap)
library(USAboundaries)
library(rgeos)
library(rgdal)
library(areal)
library(tidycensus)
```

### Data and Methods

Tornado data:

Download the tornado data from the Storm Prediction Center (SPC) http://www.spc.noaa.gov/gis/svrgis/ and load the shapefile into R.
```{r}
download.file(url = "http://www.spc.noaa.gov/gis/svrgis/zipped/1950-2017-torn-aspath.zip",
              destfile = "tornado2017.zip")
unzip("tornado2017.zip")
download.file(url = "http://www.spc.noaa.gov/gis/svrgis/zipped/1950-2017-torn-initpoint.zip",
              destfile = "tornado2017.zip")
unzip("tornado2017.zip")
TornL.sf <- read_sf(dsn = "1950-2017-torn-aspath",
                   stringsAsFactors = FALSE)
TornP.sf <- read_sf(dsn = "1950-2017-torn-initpoint",
                   stringsAsFactors = FALSE)
```

The Paths data set has missing geometries, while the points data set does not.
```{r}
any(is.na(st_dimension(TornL.sf)))
any(is.na(st_dimension(TornP.sf)))
```

Merge the two data sets. Insert point geometries where there are missing linestring geometries.
```{r}
Torn.sf <- TornL.sf
eg <- which(st_is_empty(Torn.sf))
Torn.sf$geometry[eg] <- TornP.sf$geometry[eg]
```

For missing EF ratings use the modification rules (if/else) defined here: https://www.spc.noaa.gov/wcm/OneTor_F-scale-modifications.pdf
```{r}
Torn.sf <- Torn.sf %>%
  mutate(mag = ifelse(mag == -9 & len <= 5, 0, mag),
         mag = ifelse(mag == -9 & len > 5, 1, mag))
```

Add a data/time column also add columns for path length, width, and area in metric units. Leave the time zone as native CDT.
```{r}
Torn.sf <- Torn.sf %>%
  mutate(dy = format(as.Date(date, format="%Y-%m-%d"), "%d"),
         DateTime = as.POSIXct(paste(yr, mo, dy, time), format = "%Y%m%d%H:%M:%S"),
         Hour = hour(DateTime),
         Year = year(DateTime),
         Length = len * 1609.34,
         Length = ifelse(Length == 0, min(Length[Length > 0]), Length), #takes care of zero length
         Width = wid * .9144,
         Width = ifelse(Width == 0, min(Width[Width > 0]), Width), #takes care of zero width
         Width = ifelse(Year >= 1995, Width * pi/4, Width), #takes care of change: avg to max
         cas = inj + fat,
         AreaPath = Length * Width,
         Ma = factor(month.abb[mo], levels = month.abb[1:12])) %>%
  sf::st_sf()
max(Torn.sf$yr)
```

Add energy dissipation per tornado.
```{r}
perc <- c(1, 0, 0, 0, 0, 0, 
          .772, .228, 0, 0, 0, 0,
          .616, .268, .115, 0, 0, 0,
          .529, .271, .133, .067, 0, 0,
          .543, .238, .131, .056, .032, 0,
          .538, .223, .119, .07, .033, .017)
percM <- matrix(perc, ncol = 6, byrow = TRUE)
threshW <- c(29.06, 38.45, 49.62, 60.8, 74.21, 89.41)
midptW <- c(diff(threshW)/2 + threshW[-length(threshW)], threshW[length(threshW)] + 7.5)
ef <- Torn.sf$mag + 1
EW3 <- numeric()
for(i in 1:length(ef)) EW3[i] = midptW^3 %*% percM[ef[i], ]
Torn.sf <- Torn.sf %>%
  mutate(ED = EW3 * AreaPath)
```

Isolate the Xenia, OH tornado of April 3, 1974
```{r}
Xenia.df = Torn.sf %>%
  filter(st == "OH") %>%
  filter(mag == 5) %>%
  filter(date == "1974-04-03")

print(Xenia.df$om)
print(Xenia.df$ED)
print(Xenia.df$cas)

```

Compare with historical record. How strong was the Xenia tornado.
```{r}
(length(Torn.sf$ED[Torn.sf$ED >= 2.819194e+12])/nrow(Torn.sf)) * 100

(length(Torn.sf$ED[Torn.sf$cas >= 1186])/nrow(Torn.sf)) * 100

```

The Xenia tornado is in the 99th percentile for tornado energy (99.39%) and among the top 5 tornadoes for total casualties since 1950.

Play with some graphs.
```{r}
Ohio.df = Torn.sf %>%
  filter(st == "OH")

ggplot(Ohio.df, aes(x = ED, y = cas)) +
  geom_point() + 
  scale_x_log10() + scale_y_log10()

```



