---
title: "Xenia Tornado"
author: "Tyler Fricker"
date: "2/19/2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

Set working directory and load packages.
```{r}
library(tidyverse)
library(lubridate)
library(sf)
library(tmap)
library(USAboundaries)
library(rgeos)
library(rgdal)
library(areal)
library(tidycensus)
library(scales)
```

### Data and Methods

Tornado data:

Download the tornado data from the Storm Prediction Center (SPC) http://www.spc.noaa.gov/gis/svrgis/ and load the shapefile into R.
```{r}
download.file(url = "http://www.spc.noaa.gov/gis/svrgis/zipped/1950-2017-torn-aspath.zip",
              destfile = "tornado2017.zip")
unzip("tornado2017.zip")
download.file(url = "http://www.spc.noaa.gov/gis/svrgis/zipped/1950-2017-torn-initpoint.zip",
              destfile = "tornado2017.zip")
unzip("tornado2017.zip")
TornL.sf <- read_sf(dsn = "1950-2017-torn-aspath",
                   stringsAsFactors = FALSE)
TornP.sf <- read_sf(dsn = "1950-2017-torn-initpoint",
                   stringsAsFactors = FALSE)
```

The Paths data set has missing geometries, while the points data set does not.
```{r}
any(is.na(st_dimension(TornL.sf)))
any(is.na(st_dimension(TornP.sf)))
```

Merge the two data sets. Insert point geometries where there are missing linestring geometries.
```{r}
Torn.sf <- TornL.sf
eg <- which(st_is_empty(Torn.sf))
Torn.sf$geometry[eg] <- TornP.sf$geometry[eg]
```

For missing EF ratings use the modification rules (if/else) defined here: https://www.spc.noaa.gov/wcm/OneTor_F-scale-modifications.pdf
```{r}
Torn.sf <- Torn.sf %>%
  mutate(mag = ifelse(mag == -9 & len <= 5, 0, mag),
         mag = ifelse(mag == -9 & len > 5, 1, mag))
```

Add a data/time column also add columns for path length, width, and area in metric units. Leave the time zone as native CDT.
```{r}
Torn.sf <- Torn.sf %>%
  mutate(dy = format(as.Date(date, format="%Y-%m-%d"), "%d"),
         DateTime = as.POSIXct(paste(yr, mo, dy, time), format = "%Y%m%d%H:%M:%S"),
         Hour = hour(DateTime),
         Year = year(DateTime),
         Length = len * 1609.34,
         Length = ifelse(Length == 0, min(Length[Length > 0]), Length), #takes care of zero length
         Width = wid * .9144,
         Width = ifelse(Width == 0, min(Width[Width > 0]), Width), #takes care of zero width
         Width = ifelse(Year >= 1995, Width * pi/4, Width), #takes care of change: avg to max
         cas = inj + fat,
         AreaPath = Length * Width,
         Ma = factor(month.abb[mo], levels = month.abb[1:12])) %>%
  sf::st_sf()
max(Torn.sf$yr)
```

Add energy dissipation per tornado.
```{r}
perc <- c(1, 0, 0, 0, 0, 0, 
          .772, .228, 0, 0, 0, 0,
          .616, .268, .115, 0, 0, 0,
          .529, .271, .133, .067, 0, 0,
          .543, .238, .131, .056, .032, 0,
          .538, .223, .119, .07, .033, .017)
percM <- matrix(perc, ncol = 6, byrow = TRUE)
threshW <- c(29.06, 38.45, 49.62, 60.8, 74.21, 89.41)
midptW <- c(diff(threshW)/2 + threshW[-length(threshW)], threshW[length(threshW)] + 7.5)
ef <- Torn.sf$mag + 1
EW3 <- numeric()
for(i in 1:length(ef)) EW3[i] = midptW^3 %*% percM[ef[i], ]
Torn.sf <- Torn.sf %>%
  mutate(ED = EW3 * AreaPath)
```

Add tornado paths by buffering tornado tracks by associated widths.

To perform geocomputions we need to set coordinate reference system. Here it is geographic. We transform the geographic coordinate reference system to a specific Lambert conic conformal projection.
```{r}
Torn.sf <- st_transform(Torn.sf, 
                     crs = "+proj=lcc +lat_1=60 +lat_2=30 +lon_0=-90 +units=m")
```

Buffer the geometries.
```{r}
TornB.sf <- st_buffer(Torn.sf, 
                   dist = Torn.sf$Width/2,
                   endCapStyle = 'ROUND')
```


                   
Isolate the Xenia, OH tornado of April 3, 1974
```{r}
Xenia.df = TornB.sf %>%
  filter(st == "OH") %>%
  filter(mag == 5) %>%
  filter(date == "1974-04-03")

print(Xenia.df$om)
print(Xenia.df$ED)
print(Xenia.df$cas)

```

Compare with historical record. How strong was the Xenia tornado.
```{r}
(length(TornB.sf$ED[TornB.sf$ED >= 2.819194e+12])/nrow(TornB.sf)) * 100

(length(TornB.sf$ED[TornB.sf$cas >= 1186])/nrow(TornB.sf)) * 100

```

The Xenia tornado is in the 99th percentile for tornado energy (99.39%) and among the top 5 tornadoes for total casualties since 1950.

Play with some graphs.
```{r}
Ohio.df = TornB.sf %>%
  filter(st == "OH")

ggplot(Ohio.df[Ohio.df$cas >0,], aes(x = ED, y = cas)) +
  geom_point() + 
  scale_x_log10(breaks = trans_breaks("log10", function(x) 10^x),
                     labels = trans_format("log10", math_format(10^.x))) +
  scale_y_log10() +
  xlab("Energy Dissipation [W]") +
  ylab("Number of Casualties") + 
  geom_label(label = "Xenia", 
             data = Ohio.df[Ohio.df$om == 116 & Ohio.df$Year == 1974,],
             hjust = 1.15) +
  theme_minimal()

```

Create a map
```{r}
sts <- state.name[state.name == "Ohio"]
stateBorders <- us_states(states = sts)

df = Ohio.df[Ohio.df$cas > 0,]

Xenia.df$name = "Xenia"

tm_shape(stateBorders, projection ="+init=epsg:4326") +
  tm_borders() +
  tm_fill(col = "grey94") +
  tm_shape(df) +
  tm_bubbles(col = "#fd8d3c", size = "cas", sizes.legend = c(50, 250, 500, 750, 1200), scale = 3, title.size="Casualties", alpha = 0.9) +
  tm_shape(Xenia.df) +
  tm_text(text = "name", xmod = 2.95, fontface = "bold", size = 1.5) + 
  tm_format('World', legend.position = c("RIGHT", "BOTTOM"),
                   attr.position = c("right", "bottom"),
                  legend.frame = FALSE) +
  tm_layout(frame = FALSE, attr.outside=TRUE)
```
```

test.


